---
title: "NyeriRebels Stats Analysis"
author: "Keith Karani"
format: html
editor: visual
---

### Introduction

This dashboard provides an analysis of the Nyeri Rebels Men’s Softball team statistics from the JKUAT Series which is part of the Softball Kenya Federation League 2024/2025 season. The goal is to visualize player and team performance using key metrics such as batting averages, on-base plus slugging (OPS), earned run average (ERA), and fielding performance. The dataset is sourced from the official league scoring app.

### Data loading and cleaning

```{r echo=TRUE}
library(tidyverse)
library(here)
library(janitor)
library(gganimate)


data <- read_csv(here("data", "Dekut,MenStatsFall2024.csv"))

head(data)
View(data)
```

### **Exploratory Data Analysis (EDA)**

-   Summary statistics

-   Key insights from the data

```{r}
# remove the glossary row 
data <- data[-15,]


# convert the first row into column names
data_clean <- data |> 
  janitor::row_to_names(row_number = 1)


# deduplicated column names
names(data_clean) <- make.unique(names(data_clean), sep = "_")

names(data_clean)

```

Convert the relevant column to numeric

```{r}

# convert all columns to except the first three which represent the first and last name of the players

rebels_cln <- data_clean |> 
  mutate(
    across(4:last_col(),
           ~ as.numeric(.x))
  )


rebels_cln[13, "Team"] <- "Nyeri_Rebels" 

# Replace First Name and Last Name in the 13th row with the team name



# Combine into a Single "Team/Player" Column and relocate the player name to be the first column
rebels_combined <- rebels_cln %>%
  mutate(
    player_name = if_else(
      row_number() == 13,
      "Nyeri Rebels",
      paste(First, Last)
    )
  ) %>%
  select(-Number, -First, -Last) |> 
  relocate(player_name)


view(rebels_combined)
```

### Batting performance Analysis

-   Batting Average

    One of the oldest and most universal tools to measure a hitter's success at the plate, batting average is determined by dividing a player's hits by his total at-bats for a number between zero (shown as .000) and one (1.000)

    ```{r}

    # sort players by avg

    rebels_sorted <- rebels_combined |> 
      arrange(desc(AVG))


    # Batting average
    ggplot(rebels_combined |> filter(row_number() !=13),
      aes(x = AVG, y = reorder(player_name, AVG))) +
      geom_col(fill = "steelblue") +
      labs(
        title = "Players batting average",
        x = "Batting average (AVG)",
        y = " Player Name",
        caption = "@DiamondDigest"
      ) +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 8))


    ```

-   On-base plus slugging(OPS)

    OPS adds on-base percentage and slugging percentage to get one number that unites the two. It's meant to combine how well a hitter can reach base, with how well he can hit for average and for power.

```{r}





```

-   Hits vs At-Bats

### Pitching Performance Analysis

-   Earned Run Average (ERA) Comparison

-   Strikeouts vs. Walks

### Base running analysis

-   stolen base success

### Fielding performace Analysis

-   Fielding errors vs assists

### Data exports and future work

-   saving cleaned data
-   trend analysis, machine learning predictions
