---
title: "NyeriRebels Stats Analysis"
author: "Keith Karani"
format: html
editor: visual
---

### Introduction

This dashboard provides an analysis of the Nyeri Rebels Men’s Softball team statistics from the JKUAT Series which is part of the Softball Kenya Federation League 2024/2025 season. The goal is to visualize player and team performance using key metrics such as batting averages, on-base plus slugging (OPS), earned run average (ERA), and fielding performance. The dataset is sourced from the official league scoring app.

### Data loading and cleaning

```{r echo=TRUE}
library(tidyverse)
library(here)
library(janitor)


data <- read_csv(here("data", "Dekut,MenStatsFall2024.csv"))

head(data)
View(data)
```

### **Exploratory Data Analysis (EDA)**

-   Summary statistics

-   Key insights from the data

```{r}
# remove the glossary row 
data <- data[-15,]


# convert the first row into column names
data_clean <- data |> 
  janitor::row_to_names(row_number = 1)


# deduplicated column names
names(data_clean) <- make.unique(names(data_clean), sep = "_")

names(data_clean)

```

Convert the relevant column to numeric

```{r}

# convert all columns to except the first three which represent the first and last name of the players

rebels_cln <- data_clean |> 
  mutate(
    across(4:last_col(),
           ~ as.numeric(.x))
  )


rebels_cln[13, "Team"] <- "Nyeri_Rebels" 

# Combine into a Single "Team/Player" Column and relocate the player name to be the first column
rebels_combined <- rebels_cln %>%
  mutate(
    player_name = if_else(
      row_number() == 13,
      "Nyeri Rebels",
      paste(First, Last)
    )
  ) %>%
  select(-Number, -First, -Last) |> 
  relocate(player_name)


view(rebels_combined)
```

### Batting performance Analysis

-   Batting Average

    One of the oldest and most universal tools to measure a hitter's success at the plate, batting average is determined by dividing a player's hits by his total at-bats for a number between zero (shown as .000) and one (1.000)

    ```{r}

    # sort players by avg

    rebels_sorted <- rebels_combined |> 
      arrange(desc(AVG))


    # Batting average
    ggplot(rebels_combined |> filter(row_number() !=13),
      aes(x = AVG, y = reorder(player_name, AVG))) +
      geom_col(fill = "steelblue") +
      labs(
        title = "Players batting average",
        x = "Batting average (AVG)",
        y = " Player Name",
        caption = "@DiamondDigest"
      ) +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 8))

    ```

-   On-base plus slugging(OPS)

    OPS adds on-base percentage and slugging percentage to get one number that unites the two. It's meant to combine how well a hitter can reach base, with how well he can hit for average and for power.

```{r}

# seperate the team and player data
team_ops <- rebels_combined |> 
  filter(player_name == "Nyeri Rebels") |> 
  pull(OPS)
  

# exclude team row
player_data <- rebels_combined |> 
  filter(player_name != "Nyeri Rebels")


# plot
ggplot(player_data, aes(x = reorder(player_name, OPS), y = OPS)) +
  geom_col(aes(fill = OPS > team_ops), width = 0.7) +
  #Team OPS
  geom_hline(
    yintercept = team_ops,
    color = "#BD3039",
    linewidth = 1.5,
    linetype = "dashed"
  ) +
  
  # Team OPS Label
  annotate(
    "text",
    x = 1,
    y = team_ops + 0.02,
    label = paste("Team OPS:", round(team_ops, 3)),
    color = "#BD3039",
    hjust = 0,
    fontface = "bold"
  ) +
   # Baseball-themed styling
  scale_fill_manual(
    values = c("FALSE" = "#0C2340", "TRUE" = "#FFD700"),  # Team colors
    labels = c("Below Team", "Above Team")
  ) +
  labs(
    title = "Player OPS vs. Team OPS",
    x = "Player Name",
    y = "On-Base Plus Slugging (OPS)",
    caption = "@DiamondDigest",
    fill = "Performance"
  ) +
  coord_flip() +  # Horizontal bars
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.y = element_text(size = 10),
    legend.position = "top"
  )

```

-   Hits vs At-Bats

    A hit occurs when a batter strikes the baseball into fair territory and reaches base without doing so via an error or a fielder's choice. There are four types of hits in baseball: singles, doubles, triples and home runs. All four are counted equally when deciphering batting average. If a player is thrown out attempting to take an extra base ( eg.. turning a single into a double), that still counts as a hit.

    An official at-bat comes when a batter reaches base via a fielder's choice, hit or an error (not including catcher's interference) or when a batter is put out on a non-sacrifice. (Whereas a plate appearance refers to each completed turn batting, regardless of the result.)

```{r}

player_data <- rebels_combined |> 
  filter(player_name != "Nyeri Rebels") |> 
  select(player_name, H, AB) |> 
  pivot_longer(
    cols = c(H, AB),
    names_to = "Metric",
    values_to = "Value"
  )

# grouped bar chart
ggplot(player_data, aes(x = reorder(player_name, Value), y = Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("H" = "#FFD700", "AB" = "steelblue")) +
  labs(
    title = "Hits vs At-Bats by PLayer",
    x = "Player",
    y = "Count",
    fill = "Metric",
    caption = "@Diamond Digest"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.y = element_text(size = 10),
    legend.position = "bottom")
  
```

-   On-base Percentage (OBP)

    OBP refers to how frequently a batter reaches base per plate appearance. Times on base include hits, walks and hit-by-pitches, but do not include errors, times reached on a fielder's choice or a dropped third strike. (Separately, sacrifice bunts are removed from the equation entirely, because it is rarely a hitter's decision to sacrifice himself, but rather a manager's choice as part of an in-game strategy.)

    ```{r}




    ```

### Pitching Performance Analysis

-   Earned Run Average (ERA) Comparison

-   Strikeouts vs. Walks

### Base running analysis

-   stolen base success

### Fielding performace Analysis

-   Fielding errors vs assists

### Data exports and future work

-   saving cleaned data
-   trend analysis, machine learning predictions
