---
title: "NyeriRebels Stats Analysis"
author: "Keith Karani"
format: html
editor: visual
---

### Introduction

This dashboard provides an analysis of the Nyeri Rebels Men’s Softball team statistics from the JKUAT Series which is part of the Softball Kenya Federation League 2024/2025 season. The goal is to visualize player and team performance using key metrics such as batting averages, on-base plus slugging (OPS), earned run average (ERA), and fielding performance. The dataset is sourced from the official league scoring app.

### Data loading and cleaning

```{r echo=TRUE}
library(tidyverse)
library(here)


data <- read_csv(here("data", "Dekut,MenStatsFall2024.csv"))

head(data)
View(data)
```

### **Exploratory Data Analysis (EDA)**

-   Summary statistics

-   Key insights from the data

```{r}
# Assign the 15th row as column names
colnames(data) <- as.character(data[15,])

# remove the glossary row 
data <- data[-15,]


# Resets row indices after removing the row.
rownames(data) <- NULL


glimpse(data)
str(data)
View(data)

# Remove the first row from the data 
data <- data[-1, ]

# rename the first column name to "jersey_number"

colnames(data)[1] <- "jersey_number"

# renaming check
colnames(data)

# drop the jersey_number column to remain with only relevant columns to use
data_clean <- data[,-1]

colnames(data_clean)
  
# View clean data

glimpse(data_clean)

```

Convert the relevant column to numerics

```{r}

# convert all columns to except the first two which represent the first and last name of the players

data_clean |>
  mutate(across(-c(1,2)), as.numeric)

# check any duplicate names in the columns
anyDuplicated(colnames(data_clean))

str(data_clean)

View(data_clean)

```

### Batting performance Analysis

-   Batting Average

    One of the oldest and most universal tools to measure a hitter's success at the plate, batting average is determined by dividing a player's hits by his total at-bats for a number between zero (shown as .000) and one (1.000)

    ```{r}



    ```

-   On-base plus slugging(OPS)

-   Hits vs At-Bats

### Pitching Performance Analysis

-   Earned Run Average (ERA) Comparison

-   Strikeouts vs. Walks

### Base running analysis

-   stolen base success

### Fielding performace Analysis

-   Fielding errors vs assists

### Data exports and future work

-   saving cleaned data
-   trend analysis, machine learning predictions
